---
title: "dplyr_handson"
author: 
date: "Thursday, August 14, 2014"
output: word_document
---

Start by reading in the richPK dataset or loading it from `PKPDdatasets` (sd_oral_richpk)

```{r}
rich_pk <- read.csv("datasets//richPK.csv")
```

Using dplyr 'verbs' and `%>%` to chain commands together
```{r}
library(dplyr)
filter(rich_pk, ID==1)
```


1) create a new 'demographic' dataset by subsetting out only one observation per id 

```{r}
demog_rich_pk <- rich_pk %>% filter(!duplicated(ID)) %>% select(-Time,-Amt, -Conc)
head(demog_rich_pk, n = 10)
```



  * `hint: !duplicated`
2) give basic data summary for all individuals using `summary`

```{r}
demog_rich_pk %>% summary()
```

3) summarize the demographic dataset by:
 * mean WT and AGE by gender
 
```{r}
names(demog_rich_pk) <- toupper(names(demog_rich_pk))
demog_rich_pk %>% group_by(GENDER) %>% 
  summarize(meanWT = mean(WEIGHT),
            meanAGE = mean(AGE))
```

 * range of ages by RACE
 
```{r}
demog_rich_pk %>% group_by(RACE) %>% summarize(minAGE = min(AGE),
                                               maxAGE = max(AGE))
```
4) Using the original dataset add a cmax column


```{r}
rich_pk_cmax <- rich_pk %>% group_by(ID) %>% mutate(cmax = max(Conc))
rich_pk_cmax %>% filter(!duplicated(ID)) %>% head(n = 10)
```


5) Add a HIGHCMAX column that flags all individuals over the 90th percentile
quantile

```{r}

## wrong way!! notice all HIGHCMAX values are flagged the same
wrong <- rich_pk %>% group_by(ID) %>% mutate(cmax = max(Conc), HIGHCMAX = ifelse(cmax > quantile(cmax, 0.9), 1, 0))

wrong %>% filter(!duplicated(ID)) %>% head(n = 10)

# we need to ungroup the grouping by ID before calculating the quantile!
cmax <-rich_pk %>% group_by(ID) %>% mutate(cmax = max(Conc)) %>% ungroup() %>% mutate(HIGHCMAX = ifelse(cmax > quantile(cmax, 0.9), 1, 0))

cmax %>% filter(!duplicated(ID)) %>% head(n = 10)

# check to see the proportion individuals flagged
cmax %>% filter(!duplicated(ID)) %>% summarize(proportionabove = mean(HIGHCMAX))
```


6) create an ISASIAN with a numeric (1/0) flag

```{r}
isasian <- rich_pk %>% mutate(ISASIAN = ifelse(Race == "Asian", 1, 0))
isasian %>% filter(!duplicated(ID)) %>% head(n = 10)
```

7) BONUS: Create a numeric race column where caucasian = 1, Asian = 2, Black = 3, and hispanic and other are merged into 4

```{r}
raceflg <- rich_pk %>% mutate(RACENM = ifelse(Race == "Caucasian", 1, 
                                   ifelse(Race == "Asian", 2, 
                                          ifelse(Race == "Black", 3, 4))))
raceflg %>% filter(!duplicated(ID)) %>% head(n = 10)
```

