---
title: "ggplot2 handson"
author: "Devin"
date: "Thursday, August 14, 2014"
output: word_document
---

## XY Plots

```{r}
rich_data <- read.csv("datasets//richPK.csv")
names(rich_data) <- toupper(names(rich_data))
library(ggplot2)
library(dplyr)
library(PKPDmisc)

mean_ct <- rich_data %>% group_by(TIME) %>% summarize(CONC = mean(CONC)) 



p_conc_time <- ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID)) + 
  geom_line() + 
  geom_point(size=2.5) + 
  geom_line(data = mean_ct, 
            aes(x = TIME, y = CONC, group=NULL), 
            size = 2,  color = "red") + 
  xlab("Time (hours)") +
  ylab("Concentration (mg/L)")

# regular scale
p_conc_time

# log scale
p_conc_time + scale_y_log10()

p_conc_time + facet_wrap(~RACE) + base_theme_obs()

```



2) Basic Concentration-Time plot (point and lines)
```{r}
ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID)) + 
  geom_line() + 
  geom_point()
```

3) make points/lines bigger
```{r}

ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID)) + 
  geom_line(size = 1.25) + # add size
  geom_point(size = 3)  # add size control
```

4) add log transformation to y axis
```{r}

ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID)) + 
  geom_line(size = 1.25) + 
  geom_point(size = 3) + 
  scale_y_log10() # add scale
```
5) Add color by gender
```{r}
ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID, color = GENDER)) + 
  geom_line(size = 1.25) + 
  geom_point(size = 3) 
```

6) facet by race
```{r}
ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID, color = GENDER)) + 
  geom_line(size = 1.25) + 
  geom_point(size = 3) +
  facet_wrap(~RACE) # use facet_wrap or facet_grid
```

7) add mean conc-time profile and color red
```{r}

p_conc_time <- ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID)) + 
  geom_line() + 
  geom_point(size=2.5) 

# calculate mean concentration time profile
mean_ct <- rich_data %>% group_by(TIME) %>% summarize(CONC = mean(CONC)) 

p_conc_time + 
  # add new line and specify new data to draw this line
  geom_line(data = mean_ct, 
            #don't forget group=NULL since will be looking to group by ID due to original specification in the base plot
            aes(x = TIME, y = CONC, group=NULL), 
            size = 2,  color = "red") + 
  # add some better label names too
  xlab("Time (hours)") +
  ylab("Concentration (mg/L)")

```

8) add two mean profiles (one for each gender)
```{r}
mean_ct_gender <- rich_data %>% group_by(TIME, GENDER) %>% summarize(CONC = mean(CONC)) 

p_conc_time + 
  # add new line and specify new data to draw this line
  geom_line(data = mean_ct_gender, 
            # remember to move color into aes since you're specifying
            #a column in the dataset to specify the coloring
            aes(x = TIME, y = CONC, group=NULL,  color = GENDER), 
            size = 2) + 
  # add some better label names too
  xlab("Time (hours)") +
  ylab("Concentration (mg/L)")
```

9) color by weight
```{r}
ggplot(data = rich_data, 
                      aes(x = TIME, y = CONC, group=ID, color = WEIGHT)) + 
  geom_line(size = 1.25) + 
  geom_point(size = 3)
```

## Boxplots and Histograms
* Histogram(s) of demographics
```{r}
# filter to one observation per ID
ggplot(data = rich_data %>% filter(!duplicated(ID)), 
       aes(x = WEIGHT)) + 
  geom_histogram()

# clean up with customizations in geom_histogram
ggplot(data = rich_data %>% filter(!duplicated(ID)), 
       aes(x = WEIGHT)) + 
  geom_histogram(binwidth= 4, color="black", fill="white")
```


  * add vertical line for median value
```{r}
median <- rich_data %>% filter(!duplicated(ID)) %>% summarize(medianWT = median(WEIGHT))
ggplot(data = rich_data %>% filter(!duplicated(ID)), 
       aes(x = WEIGHT)) + 
  geom_histogram(binwidth= 4, color="black", fill="white") +
  geom_vline(xintercept = median[["medianWT"]], size= 2, color = "red")
```

* Boxplot cmax/AUC by gender
```{r}
cmaxauc <- rich_data %>% group_by(ID, GENDER) %>% summarize(cmax = max(CONC), aucinf = AUC_inf(TIME, CONC))
ggplot(cmaxauc, aes(x = GENDER, y = cmax, group = GENDER)) + geom_boxplot()
ggplot(cmaxauc, aes(x = GENDER, y = aucinf, group = GENDER)) + geom_boxplot()
```

## Eta-Eta plots
Read in `EtaCov_base` dataset
```{r}
etacov_base <- read.csv("datasets/EtaCov_base.csv")
```

* Eta-Eta pairs plot
```{r}
library(GGally)
ggpairs(etacov_base %>% select(nV, nCl, nKa),
        lower=list(continuous="smooth", params=c(colour="black")),
        diag=list(continuous="density", params=c(colour="black", fill = "white")), axisLabels = "show")

```

* Eta-cov scatter plot
```{r}

etacov_base$Scenario <- NULL
library(reshape2)
enc_melt <- melt(etacov_base, id.vars=c("ID", "WT"))
gg_covwt <- ggplot(enc_melt, aes(x = WT, y=value, group = ID)) 
gg_covwt+ geom_point() + 
  facet_wrap(~variable)

```

* add loess fit
```{r}
gg_covwt+ geom_point() + 
  facet_wrap(~variable)+ stat_smooth(aes(group= variable), se=F, size = 1.25)
```

