---
title: "Grouping, Summary Statistics, and Reshaping Data"
author: "Devin"
date: "Thursday, August 14, 2014"
output: word_document
---

---
## One (summarization) package to rule them all?
* built in functions like aggregate and the old (but still useful) package plyr make group summary statistics easy
* **dplyr** makes everything a breeze

---
## Key dplyr Functions
* filter: keeps rows with matching criteria
* summarize: do calculations *and* reduce resulting variables and values
* mutate: do calculations *and* add them to existing data frame
* arrange: ordering
* select: picking columns by name

#### wrapped with `%>%` and tied with a `group_by` bow


---
# What does is look like

```{r}
library(dplyr)
data %>% 
  group_by(ID) %>% 
    summarize(cmax = max(CONC)) %>% 
      head

```

## Go to dplyr handson

---
```{r, include =F}
library(PKPDdatasets)
data <- sd_oral_richpk
names(data) <- toupper(names(data))
```

# Reshaping Data

There is an excellent package (library) called reshape2

```{r}
library(reshape2)
dat_10IDS <- data %>% filter(ID < 10)
m_dat_10IDS <- melt(dat_10IDS %>% select(ID,TIME, AMT, CONC, AGE, WEIGHT, DOSE), id.vars="ID")
head(m_dat_10IDS)
```



---
## What about long to wide
How can we make a table of concentrations by time for all ID's? (long to wide conversion)

---
```{r}
molten <- melt(data %>% filter(ID < 6), id.vars = c("TIME", "ID"), measure.vars = c("CONC"))
head(molten)
tail(molten)
concTimeWide <- dcast(molten, TIME + variable ~ ID)
head(concTimeWide, n = 4)
```
