---
title: "Merging and Sorting"
author: "Devin"
date: "Thursday, August 14, 2014"
output: word_document
---
```{r}
library(PKPDdatasets)
data <- sd_oral_richpk
names(data) <- toupper(names(data))
```


---
## Create some Data to Merge
```{r}
tableOne <- data.frame(ID = c(1:3), PERIOD = c("a", "b", "c"))
tableTwo <- data.frame(ID = c(seq(1,5,2)), PERIOD= c("a", "c", "f"),ISM = c(0, 1, 0) )

```
```{r}
tableOne
tableTwo
```

---
## Merging Options
Can merge by similarities in specific column, specific index, or can combine all
```{r}
merge(tableOne, tableTwo, by = "ID")
merge(tableOne, tableTwo, by = "ID", all = T)
```

---
## Merging Options (II)
```{r}
merge(tableOne, tableTwo, by = "ID", all.x = T)
merge(tableOne, tableTwo, by = "ID", all.y = T)
```

---
## Merging Options (III)
Can also bind together next to one another
```{r}
cbind(tableOne, tableTwo)
```

---
## Merging Options (IV)

Can stack two data frames using `rbind` but data frames must have equal number of columns

```{r}
tableOne <- data.frame(ID = c(1:3), PERIOD = c("a", "b", "c"))
tableTwo <- data.frame(ID = c(seq(1,5,2)), PERIOD= c("a", "c", "f"))
rbind(tableOne, tableTwo)
```

---


---
## Sorting Data

Lets sort the data by time

```{r}
keep <- c("CID", "TIME", "CONC", "DOSE")
sorted <- data[, keep]
sorted <- sorted[order(sorted$TIME),]
head(sorted)
```

--- 
## Multiple Sorts at Once

First let's 'mess up' the dataset - sort by descending concentrations
```{r}
concSort <- sort[order(-sort$CONC),]
head(concSort)
```

--- 
##Mutliple Sorts at Once (II)
- let's see what happens if we sort by ID
```{r}
concSort <- concSort[order(concSort$CID),]
head(concSort)
```

Oh no the times aren't in order!

---
## Multiple Sorts at Once (III)

Thankfully we can fix this by sorting ID **and** time within ID 
```{r}
concSort <- concSort[order(concSort$CID, concSort$TIME),]
head(concSort)
```

---
## Sample Dosing Sheet Creation

```{r}
conc <- c(0)
nSubj <- 40
times <- c(0, 0.5, 1, 2, 3, 4, 6, 8, 12, 16, 24)
t <- expand.grid(TIME = times, CID = 1:nSubj, CONC = conc)
t <- as.data.frame(t)
democols <- c("CID", "AGE", "WT", "CLCR")
demo <- unique(data[, democols])
res <- merge(t, demo, by = "CID")
head(res)
```
